Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2018-12-05 Manuscript/manuscript.Rmd
manuscript edits

---
title: "Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings"
csl: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/styles.ref/genomebiology.csl
output: 
  word_document:
    #reference_docx: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/styles.doc/NIH_grant_style.docx
bibliography: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/References/references.bib
---

@@

The mechanisms underlying the formation of TADs are a complex and active area of research. Recently, it has been discovered that insulator sequences have a primary role in orchestrating the topological arrangements of higher-order chromatin architecture [@phillips2013chromatin]. Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long-range chromatin contacts [@zuin2014cohesin]. Other regulatory elements like particular histone modifications such as H3k36 and H3K27 trimethylation, and transcription factors such as ZNF274 and YY1 have been more frequently observed at TAD boundaries than in other regions [@rao20143d]. These have been found to be associated with more open chromatin which alter the accessibility of genes for transcription. The distinct patterns of some of these different proteins and functional elements point toward the opportunity of using computational approaches in identifying which epigenomic features are most predictive of the development of TAD boundaries. This will allow us to better understand what leads to their formation.

Most accepted methods for predicting TAD boundaries consider performing a classification algorithm by binning the genome into equal-sized bins and labeling them as either overlapping with a flanked TAD boundary or not. However, a pervasive issue becomes the introduction of highly imbalanced classes. Due to the sparcity of TADs throughout the genome, most bins will not contain a TAD boundary. Therefore, the minority class of genomic bins with TAD boundaries in them becomes proportionally smaller than that of the majority class of genomic bins with no TAD boundaries. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple].  A clear method for creating balanced classes remains evasive and largely dependent on the problem at hand.

Another problem lies in identifying the feature space of subsequent classification algorithms.  A variety of positional characteristics of the genomic features can be used to describe the relationship between TAD boundaries and genomic elements, but it is unclear which is best. The use of distances (in base pairs) from genomic bins to regions of functional genomic elements has largely been ignored. However, these distance characteristics offer a more accurate spatial representation of which genomic features are associated with TAD boundaries.

Due to the increasingly large amounts of functional genomic elements to consider, it is a popular approach in genomic studies to use variable reduction techniques to reduce the noise induced by noninformative predictors and increase computational efficiency. Two popular approaches use the $l_{1}$ norm (LASSO) and a combination of the $l_{1}$ and $l_{2}$ norms together (Elastic-Net). While both regularization techniques have merit, LASSO’s inability to handle correlated predictors poses a problem when identifying which element are most predictive of TAD boundaries. The elastic-net, however, is better suited in a situation where multiple genomic elements colocalize around the same TAD boundaries, and could be better suited for this type of alaysis [@zou2005regularization].

Model performance is also contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. This will also affect the size of the genomic bins used in model construction. A larger resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances.

Careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models. A common metric to consider is the the area under the ROC curve (AUC), as well as others including accuracy, sensitivity, and specificity.

The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a genomic bin overlapping with a TAD boundary region--mean probabilities are biased towards the majority class. Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance on imbalanced data. More robust metrics like the F1-score and Matthew's Correlation Coefficient are more suitable for this type of analysis and can highlight problems associated with applying predictive models on imbalanced data.

In this study we have developed an ensemble framework that aims to address and compare a variety of data irregularities and characteristics that are associated with the prediction of TAD boundaries. We evaluated domain data obtained from four different resolution contact matrices for the GM12878 cell line including 5kb, 25kb, 50kb, and 100kb. For each resolution, we examined the inclusion of three different predictor types as the feature space for downstream modeling. We coupled each predictor type with two separate variable reduction techniques. Lastly, we performed a series of three different re-sampling techniques aimed at created balanced classes. Random forest classification algorithms were performed to both compare each combination of the above ensemble, and investigate which functional genomic features were most associated with the formation of TAD boundaries.  Model performance was evaluated based on a total of four metrics which included test accuracy, area under the ROC curve (AUC), F1 score, and Matthew's Correlation Coefficient (MCC).

@@

To establish the data for modeling, the genome was binned into equally sized intervals (Figure 18 C). This was performed at the same resolution of the contact matrix that the domain data was obtained from. For example, for the domain data obtained from a 5kb resolution contact matrix, the genome was subsequently binned into 5kb intervals. The same processes was repeated for each of the other three resolutions. Next, the boundary coordinates were flanked on either side by a unit corresponding to the resolution of the particular domain data. For example, for the domain data obtained from a 5kb resolution contact matrix, each uniquely identified TAD boundary was flanked on either side by 5kb for a total width around each boundary point of 10kb. The same process was repeated for the other three resolutions. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

Using each genomic element, we created three types of predictors that were used as features to analyze their relationships with TAD boundaries in downstream modelling. The predictors consisted of two different types of overlaps, and a distance type. They are described in detail below:

   1.	Overlap Counts (OC): For each genomic bin, the total number of overlapped regions between a functional genomic element and a bin was calculated. This was defined as the overlap counts (OC) and is illustrated in the left part of Figure 2. Within each bin (marked by the blue dashed lines), the total number of overlapped regions of the genomic element (represented by the horizontal green segments) are tallied and marked below. In the case of no overlaps the value is given as 0 (Figure 2 A).

   2.	Overlap Percent (OP): For each genomic bin, the percent overlap between regions defined by a particular functional element and a bin was calculated. From the middle diagram of Figure 2, the feature width was defined as the number of base pairs that overlapped with the genomic bin. The percent was then calculated by dividing the feature width with the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of base pairs that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 5kb, 25kb, 50kb, or 100kb). Similar to the OC predictor type, bins with no overlaps were given to be 0 (Figure 2 B).

   3.	Distance: For each genomic bin, the distance (in base pairs) from the center of the bin to the center of the nearest region of a respective functional element was calculated. The right most diagram of Figure 2 depicts an example of how distance was calculated. The vertical green segments represent the centers of the respective genomic element. Each distance, $D_{i}$, represented by the blue brackets are the number of base pairs from the center of the bins to the center of the nearest genomic elemental region (Figure 2 C).

@@

###Training and Testing Sets

@@

###Variable Reduction

@@

###Re-Sampling Techniques

@@

   2.	Random Under-Sampling: All of the minority classes from the training set were used. A random set without replacement from the majority classes of the training set were used to match the number of minority samples. In this paper we performed 50 iterations of random under-sampling. At each iteration a random forest classification algorithm was performed. Performance metrics were aggregated by taking the average across all of the iterations.

   3.	Random Over-Sampling: All of the majority classes from the training set were used. A random set with replacement from the minority classes of the training set were used to match the number of majority samples. Similar to the under-sampling technique, 50 iterations were performed and results were aggregated by taking the average across iterations.

   4.	SMOTE (Synthetic Minority Over-Sampling Technique): This method incorporated both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class. The SMOTE algorithm is stored in the DMwR package, available in R [@chawla2002smote]. The user is able to specify the parameters denoting the percent of over- and under-sampling from the two classes. This allows the user to control the proportion of disbalance in the data. For the purposes of this paper, we used SMOTE to create perfectly balanced data. However further research would be needed to see if varying the proportion of class disbalance could improve predictive performance.

###Classification

@@

###Model Evaluation

@@

$$
\begin{equation}
\begin{aligned}
Accuracy &= \dfrac{TP + TN}{TP + TN + FP + FN} \\
F1 - Score &= \dfrac{2TP}{2TP + FP + FN} \\
MCC &= \dfrac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}
\end{aligned}
\end{equation}
$$
Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set. Accuracy measures the percentage of correct classifications by the model. However, in the case of heavily imbalanced class, as can be seen by the numerator, the metric is biased toward correcly identified majority classes. The F1-score and MCC are more balanced measures. We also considered the AUC for each model, by computing the average trapezoidal approximations for the curve created by the true positive rates (TPR) and false positive rates (FPR).

@@

In this section we provide the details of the comprehensive experiments performed in order to assess each ensemble system. For the task of predicting TAD boundaries using functional genomic elements, we evaluated 4 domain data sets, 3 predictor types, 2 variable reduction techniques, and 4 class balancing techniques. We evaluated each ensemble system using 10-fold cross-validated random forest models. Thus, we generated 96 (=4 $\times$ 3 $\times$ 2 $\times$ 4) separate models. For simplicity, we have only presented the most significant results. Results of the complete ensemble are available in the Supplementary Materials.

>>>>

---
title: "Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings"
#csl: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/styles.ref/genomebiology.csl
output: 
  word_document:
    #reference_docx: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/styles.doc/NIH_grant_style.docx
bibliography: C:/Users/Spiro Stilianoudakis/Documents/TAD_predictive_modeling/Manuscript/References/references.bib
---

@@

The mechanisms underlying the formation of TADs are a complex and active area of research. Recently, it has been discovered that insulator sequences have a primary role in orchestrating the topological arrangements of higher-order chromatin architecture [@phillips2013chromatin]. Insulators are multi-faceted regulatory sequences that moderate a variety of genomic processes including activation, repression, and enhancer blocking. Specifically, the insulator binding protein CTCF has been found to be enriched at the boundary sequences of topologically associating domains in human cells and may therefore act as a mediator of long-range chromatin contacts [@zuin2014cohesin]. Other regulatory elements like particular histone modifications such as H3k36 and H3K27 trimethylation, and transcription factors such as ZNF274 and YY1 have been more frequently observed at TAD boundaries than in other regions [@rao20143d]. These have been found to be associated with more open chromatin which alter the accessibility of genes for transcription. The distinct patterns of different proteins and functional elements point toward the opportunity of using computational approaches in identifying which epigenomic features are most predictive of the development of TAD boundaries, allowing us to better understand what leads to their formation.

A study published by Mourad et al. used multiple logistic regression with LASSO regularization to predict TAD boundaries [@mourad2016computational]. They considered binning the genome at both 50 base and 1kb intervals and flanking TAD boundaries by 1kb and 20kb regions. The outcome Y was determined by whether or not a genomic bin overlapped with a flanked boundary. The feature space was obtained by considering the percent overlap between a bin and coordinates defining ChIP-seq peak data. Models were then evaluated using AUROC. However, a pervasive issue here becomes the introduction of highly imbalanced classes. Due to the sparcity of TADs throughout the genome, most bins will not overlap with a flanked TAD boundary. It is commonly agreed upon in the machine learning community that imbalanced datasets adversely impact the performance of the classifiers as the learned model is biased towards the majority class to minimize the overall error rate [@estabrooks2004multiple]. Similarly, careful consideration must also be taken when deciding which performance metric(s) to present when evaluating predictive models, especially in the case of imbalanced data. The popularity surrounding the use of ROC curves is that they do not depend on a threshold. However, when the prevalence of an event is low--that is, a genomic bin overlapping with a TAD boundary region--mean probabilities are biased towards the majority class. Likewise, metrics like accuracy and sensitivity and specificity do not incorporate the entirety of the confusion matrix created when validating predictive models on a test data set. Therefore, each of these metrics are likely to be inflated when evaluating model performance.

In a similar study, Hong et al. used domain data from an hESC cell line to build a linear model and predict which genomic elements were associated with TAD boundaries [@hong2017computational]. They defined boundary segments as genomic regions before and after 150 kilobases of the boundary. To solve the imbalance problem, they randomly sampled from regions outside of the 150kb flank.  For their model, the feature space was determined by the number of counts of ChIP-seq peaks defining functional genomic elements that overlapped with the two segments. Individual models were then performed until the best model which included the concensus CTCF signal feature, followed by 10 additional elements was obtained. Here, however, it is likely that 1 iteration of random under-sampling will yield biased results, depending on how representative the under-sampled data is. Likewise, it is unclear how the performances of other balancing techniques compare (i.e. over-sampling or a combination of both under and over-sampling).

Furthermore, there are other aspects of TAD boundary modelling that current methods have failed to consider. Most notably is the concept of distance. As a feature engineering step, the use of distances (in base pairs) from genomic bins to regions of functional genomic elements has largely been ignored. However, these distance characteristics have the potential to offer a more accurate spatial representation of which genomic features are associated with TAD boundaries. Additionally, model performance is heavily contingent upon the resolution of the contact matrix used to identify the location of TADs throughout the genome. It is often unclear from authors of previous studies what resolution contact matrices were when retrieving TAD boundaries, and how other resolutions would affect their results. A larger resolution contact matrix will result in the identification of larger-sized TADs, and subsequently, less TAD boundary points. This will contribute to smaller sample sizes and could effect predictive performances.

In this study we have developed an ensemble framework that aims to address and compare a variety of data irregularities and characteristics that are associated with the prediction of TAD boundaries. We evaluated domain data obtained from four different resolution contact matrices for the GM12878 cell line including 5kb, 25kb, 50kb, and 100kb. For each resolution, we examined the inclusion of three different predictor types as the feature space for downstream modeling. We coupled each predictor type with two separate variable reduction techniques. Lastly, we performed a series of three different re-sampling techniques aimed at creating balanced classes. Random forest classification algorithms were performed to both compare each combination of the above ensemble, and investigate which functional genomic features were most associated with the formation of TAD boundaries.  Model performance was evaluated based on a total of four metrics which included test accuracy, area under the ROC curve (AUC), F1 score, and Matthew's Correlation Coefficient (MCC).

@@

To establish the data for modeling, the genome was binned into equally sized intervals (Figure 18 C). In order for a more systematic approach, the binning process was performed at the same resolution of the contact matrix that the domain data was obtained from. For example, for the domain data obtained from a 5kb resolution contact matrix, the genome was subsequently binned into 5kb intervals. The same processes was repeated for each of the other three resolutions. Again, operating systematically, we flanked the boundary coordinates on either side by a unit corresponding to the resolution of the particular domain data. For example, for the domain data obtained from a 5kb resolution contact matrix, each uniquely identified TAD boundary was flanked on either side by 5kb for a total width around each boundary point of 10kb. The same process was repeated for the other three resolutions. A genomic bin was labeled as having a TAD boundary within it (Y=1) if it overlapped with a particular flanked boundary. Otherwise, the bin was labeled as not containing a boundary (Y=0).

@@

Using the genomic elements, we created three types of predictors that were used as features to analyze their relationships with TAD boundaries in downstream modelling. The predictors consisted of two different types of overlaps, and a distance type. They are described in detail below:

   1.	Overlap Counts (OC): For each genomic bin, the total number of instances where a bin overlapped with regions representing a functional genomic element was calculated (Figure 2 A). In the case of no overlaps the value is given as 0.

   2.	Overlap Percent (OP): For each genomic bin, the percent overlap between the feature width and the total width of the bin was calculated (Figure 2 B). We defined the feature width as the number of base pairs that overlapped with the genomic bin. The percent was then calculated by dividing the feature width by the bin width. If multiple overlaps existed, the feature width was defined as the sum of the total number of base pairs that overlapped with a particular bin. Note that the bin width remained constant for a given resolution (either 5kb, 25kb, 50kb, or 100kb). Bins with no overlaps were given to be 0.

   3.	Distance: For each genomic bin, the distance (in base pairs) from the center of the bin to the center of the nearest region of a respective functional element was calculated (Figure 2 C).

@@

###TRAINING & TESTING SETS

@@

###VARIABLE REDUCTION

@@

###RE-SAMPLING TECHNIQUES

@@

   2.	Random Under-Sampling (RUS): All of the minority classes from the training set were used. A random set without replacement from the majority classes of the training set were used to match the number of minority samples. In this paper we performed 50 iterations of random under-sampling. At each iteration a random forest classification algorithm was performed. Performance metrics were aggregated by taking the average across all of the iterations.

   3.	Random Over-Sampling (ROS): All of the majority classes from the training set were used. A random set with replacement from the minority classes of the training set were used to match the number of majority samples. Similar to the under-sampling technique, 50 iterations were performed and results were aggregated by taking the average across iterations.

   4.	SMOTE (Synthetic Minority Over-Sampling Technique): This method incorporated both random under- and over-sampling. Under-sampling is performed without replacement from the majority class, while over-sampling is performed by creating new synthetic observations using the minority class. The SMOTE algorithm is stored in the DMwR package, available in R [@chawla2002smote]. The user is able to specify the parameters denoting the percent of over- and under-sampling from the two classes. This allows the user to control the proportion of disbalance in the data. For the purposes of this paper, we used SMOTE to create perfectly balanced data.

###CLASSIFICATION

@@

###MODEL EVALUATION & PREDICTION

@@

$$
\begin{equation}
\begin{aligned}
Accuracy &= \dfrac{TP + TN}{TP + TN + FP + FN} \\
F1 - Score &= \dfrac{2TP}{2TP + FP + FN} \\
MCC &= \dfrac{TP \times TN - FP \times FN}{\sqrt{ (TP + FP)(TP+FN)(TN+FP)(TN+FN) }}
\end{aligned}
\end{equation}
$$
Here, TP refers to the number of bins correctly identified as containing a TAD boundary (true positives), FP refers to the number of bins incorrectly identified as containing a TAD boundary (false positives), TN refers to the number of bins correctly identified as not containing a TAD boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a TAD boundary. Each of these quantities are obtained from the confusion matrix created by validating the model on the testing set.  We also considered the AUC for each model by computing the average trapezoidal approximations for the curve created by the true positive rates (TPR) and false positive rates (FPR).

For the task of identifying the functional genomic elements that were most predictive of the formation of TAD boundaries, we evaluated the variable importances associated with each predictor of the random forest model. For each tree, the prediction accuracy on the out-of-bag portion of the data is recorded. Then the same is done after permuting each predictor variable. The difference between the two accuracies are then averaged over all trees, and normalized by the standard error [@kuhn2012caret].

@@

In this section we provide the details of the comprehensive experiments performed in order to assess each ensemble system. We evaluated and compared 4 domain data sets, 3 predictor types, 2 variable reduction techniques, and 4 class balancing techniques. Thus, we generated 96 (=4 $\times$ 3 $\times$ 2 $\times$ 4) separate models. For simplicity, we have only presented the most significant results. Results of the complete ensemble are available in the Supplementary Materials.
